Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-06-22 manuscript/manuscript.tex
Review of final section
Overall review of final sections to summarize content to most relevant points.

\section{Data Dictionary Observations} 
\label{sec:datadictionay}
During this analysis, it became apparent that the  Data 
Dictionary could use an update. We found several discrepancies 
between the Data Dictionary and the actual data which can lead 
analytic efforts astray.

===========================================
REVIEW BELOW HERE
===========================================

\begin{itemize}
	\item While the Open Data portal lists the basic data types of the 
	various fields, the Data Dictionary does not. Nor does it provide 
	meaningful information on the specifics of various fields, such 
	as constraints or domain of legal values. For example, the 
	incident\_zip field is specified as ``text''. That's probably not the 
	best definition, perhaps	``categorical'' would be more 
	appropriate with a description indicating that the data can 
	contain only numeric characters and is not subjected to arithmetic 
	operations. We found two instances with the zip\_code field 
	contained character strings (``na'', ``N/A'').

	\item Domain of legal values as contained in the Data Dictionary 
	are incomplete. For example, the Data Dictionary indicates that 
	the status field has values of assigned, canceled, closed, or pending. 
	However, we found additional status values: in progress, started, 
	and unspecified. Additionally, we discovered no SRs with a status of 
	canceled. Similarly, the address\_type field is indicated to have 
	values address, blockface, intersection, latlong, and placename. We 
	discovered additional values of bbl and unrecognized. No values of 
	latlong were observed. Other fields suffer the same 	inaccuracies 
	(facility\_type, vehicle\_type, taxi\_pick\_up\_location, road\_ramp, city).
\end{itemize}

\section{Recommendations} 
\label{sec:recommendations}
Having identified a number of data cleanliness related issues with the 
311 Service Request (SR) dataset, what follows are a series of 
recommendations as to how to resolve these issues. Below are 
those recommendations;

\begin{itemize}
	\item First, and most importantly, the Open Data Team will need 
	to undertake a special effort to identify these data anomalies, confirm 
	the veracity of the issue, and engage with the appropriate Agency 
	Open Data Representatives to resolve these issues. That effort will 
	not be easy, fast, or cheap as in some software changes will likely 
	be required. Perhaps hardest will be working with the external City 
	Agencies that feed data to the 311 SR dataset, but are not using 
	the core 311 system. Integrations and API usage will need to be 
	investigated, and in many cases the changes required to correct 
	an issue will be on the source Agency side. Coordination and 
	mutual approval of any changes will be key. Strong support from 
	senior 	leadership will be a critical success factor.

	\item Secondly, standards will be to be identified, agreed to, and 
	commitment to support those standards must be obtained. For 
	example, what are the domain of legal values for certain fields, 
	especially those with identified issues?

	\item Obviously, many of the data integrity issues can be 
	resolved through rule enforcement by software. For example, the 
	software could easily prevent invalid values, such as zip codes, from 
	entering the system. Similarly for an entire range of data issues. One 
	key area is the created\_date and closed\_date which are frequently 
	cited when determining responsiveness of various Agency services. The 
	authors would argue that such dates should be driven by change 
	in ``status'', not manually entered. For example, the closed\_date 
	is captured when the SR ``status'' is changed to ``closed'', and 
	not manually entered. Such a change, along with many others, could 
	prevent errors such as negative and  zero durations which 
	significantly affect measurements of responsiveness.

	\item The large spikes in SR closure and creation occurring at 
	midnight and noon, almost certainly point to an underlying automated 
	process from an external-to-311 Agency. Such processes again distort 
	the accurate capture of an SR duration. Data rule enforcement at the 
	software integration or API level should be investigated.

	\item One possible way to have software enforce certain data 
	standards, would be to move more non-311 Agencies to the core 
	311 software system. That system was completely upgraded in 
	2019, and should be (relatively) easy to expand to other 
	Agencies and usages.

	\item As noted, there are six ``computed'' fields in the data export 
	that are not identified in the Data Dictionary. While these fields may 
	be ``experimental'', they are nonetheless present in any data portal 
	presentation or export. As noted in this study, a number of those 
	computed fields suffer from data inaccuracies. It is difficult to know 
	what to recommend for these six field without further understanding of 
	their origin, computational method, or intended usage. We would 
	recommend updating the Data Dictionary at a minimum and 
	possibly hiding those fields from public display until such time as 
	their usage and accuracy can be ascertained.

	\item Removal of duplicate fields should be examined.  In some cases, such 
	as the borough and park\_borough fields, the duplication is 100\%. In 
	other cases, the duplication is in the 80-90\% range. These fields and 
	their usage/utility should be carefully examined to determine if 
	their inclusion in the 311 SR dataset is meaningful, adds value not found 
	in other fields, and should be removed or continued to be collected. Initial 
	analysis indicates that as much as 12\% size reductions are achievable 
	with the removal of duplicate fields. A more thorough review and 
	understanding of the fields would no doubt yield more savings.

	\item Fields with extremely low data presence, such as taxi\_company\_borough 
	which is populated only 0.06\% of the time. Similarly for several other 
	fields evaluated in this study which have a high percentage of blank/unknown 
	values. If the presence of a small percentage of SRs with these fields 
	being populated is deemed useful and necessary, then so be it. But if 
	the usage of these fields is so diminished, then perhaps its inclusion 
	in the 311 SR dataset is not warranted.

	\item The Data Dictionary is in need of an upgrade and update. The last 
	revision is indicated at 6 March 2023, well over a year ago. 	As noted 
	in \textbackslash nameref: \nameref{sec:datadictionary} there are many 
	problems with the Data Dictionary, including domain values, absence of 
	the six ``computed'' fields, and inaccurate description of the usage of 
	several fields. Some of the Data Dictionary entries can be undertaken 
	immediately, which further revision should be addressed as part of 
	a deeper data clean-up initiative.

	\item The excessive precision in the latitude, longitude, and location 
	fields is simply nonsensical.  We would recommend no more than 
	four decimal places (0.000X) which represents a distance of approximately 
	36.4 feet. Five decimals (0.0000X) represents 3.6 feet, which although 
	possible, would be an extremely sophisticated and accurate 
	geo-coding system. A 14 decimal number is simply a case of a lot 
	of precision with limited accuracy, and should be avoided.

	\item The 10 address-related address fields (incident\_address, 
	street\_name, city, intersection\_street\_1 \& 2, cross\_street\_1 \& 2,  
	landmark, road\_ramp, bridge\_highway\_segment, and 
	taxi\_pick\_up\_location should be in standard USPS format. There 
	are several software packages which can standardize addresses to 
	these standards. We believe such an effort should be automatically 
	applied to these fields at data entry. This step would also greatly 
	improve any underlying geocoding routines which 	may be used 
	to determine data values such as BBL, latitude, longitude, and New 
	York State X/Y-plane coordinates.

	\item The use of the value ``UNKNOWN'' , ``N/A'', ``NA'', and ``na'' when 
	entered directly as text into a field should be very carefully examined, 
	and in all likelihood prohibited.  There are some indeed a few 
	complaint\_type(s) where an ``N/A'' value is appropriate, such 
	as a consumer\_complaint about certain non-geocentric issues, such 
	as a travel agency or tax preparation service . But even in those cases 
	a value of ``N/A'' should be entered only via a drop-down menu 
	selection. During this analysis, there were far too many data 
	representations of missing data including: <N/A>, ``na'', `` ``, UNKNOWN, 
	and ``N/A''. It's challenging for analyst to have to account for so 
	the many different methods of representing empty or null data fields. 
\end{itemize}

>>>>

\section{Data Dictionary Observations} 
\label{sec:datadictionary}
During this analysis, it became apparent that the  Data 
Dictionary could use an update. We found several discrepancies 
between the Data Dictionary and the actual data which can lead 
analytic efforts astray.

\begin{itemize}
	\item While the Open Data portal lists the data types of the 
	various fields, the Data Dictionary does not. Nor does it provide 
	much information on the specifics of various fields, such 
	as constraints or domain of legal values. For example, the 
	incident\_zip field is specified as ``text''. That's probably not the 
	best definition, perhaps	``categorical'' indicating that the data can 
	contain only numeric characters and is not subjected to arithmetic 
	operations.

	\item Many of the domain of legal values are incomplete or 
	inaccurate. For example, the Data Dictionary indicates that 
	the ``status'' field has values of \textit{assigned, canceled, closed}, 
	or \textit{pending}. However, we found additional status 
	values of: \textit{in progress, started}, and \textit{unspecified}. 
	However, we found no SRs with a status of \textit{canceled}. 
	Similarly, the address\_type field is indicated to have 
	values \textit{address, blockface, intersection, latlong}, and 
	\textit{placename}. We discovered the additional values of 
	\textit{bbl} and \textit{unrecognized}. Additonally, no values of 
	\textit{latlong} were observed. Other fields (facility\_type, 
	vehicle\_type, taxi\_pick\_up\_location, road\_ramp, city) 
	suffer similar inaccuracies in their doman of values.

	\item As noted previously, there are six \textit{@computed} fields that
	are present in the data, but not described in the Data Dictionary.
\end{itemize}

\section{Recommendations} 
\label{sec:recommendations}
Here are recommendations as to how to resolve these issues.

\begin{itemize}
	\item First, and most importantly, the Open Data Team will need 
	to undertake a dedicated effort to identify these 
	data anomalies, confirm the veracity of the issue, 
	and engage with the appropriate Agency Open Data 
	Representatives to resolve these issues. That effort will not 
	be easy, fast, or cheap and in some cases software changes will likely 
	be required. Coordination and mutual approval of any 
	changes will be key, and solid support from senior leadership 
	will be required.

	\item Secondly, standards and protocols will need to be 
	identified, agreed to, and commitment obtained. For example, 
	what are the domain of legal values for certain fields?

	\item Many of the data integrity issues could potentially be 
	resolved through rule enforcement by software. For example, the 
	software could easily prevent invalid values, such as zip 
	codes. Similarly for other data issues. One key area are is the 
	created\_date and closed\_date which are frequently cited 
	when determining responsiveness of various Agency services.

	\item The large spikes in SR closure and creation occurring at 
	midnight and noon, almost certainly point to an automated 
	process from an external Agency integration. Such processes again distort 
	the accurate capture of an SR duration.

	\item Investigate moving additional Agencies to the core 
	311 software system where enforcement of data rules would 
	presumably be more consistent.

	\item As noted, there are six ``computed'' fields in the data export 
	that are not identified in the Data Dictionary. As noted, several of 
	these computed fields suffer from data inaccuracies. We would 
	recommend updating the Data Dictionary at a minimum and 
	possibly hiding those fields from public usage until such time as 
	their accuracy can be ascertained.

	\item Removal of duplicate fields should be examined.  In some cases, such 
	as the borough and park\_borough fields, the duplication is 100\%. In 
	other cases, the duplication is in the 80-90\% range. These fields and 
	their usage/utility should be carefully examined to determine if 
	their inclusion in the 311 SR dataset is meaningful and adds value not found 
	in other fields. Initial analysis indicates that as much as 12\% 
	size reductions are achievable.

	\item Fields with extremely low data presence, such as taxi\_company\_borough 
	which is populated only 0.06\% of the time, should be evaluated 
	to determine if their inclusion is warranted.

	\item The Data Dictionary is in need of an upgrade and update. The last 
	revision is indicated at 6 March 2023, well over a year ago. 	As noted 
	in Section~\ref{sec:datadictionary} there are several problems 
	with the Data Dictionary, including domain values, absence of 
	the six ``computed'' fields, and inaccurate description of the usage of 
	several fields.

	\item The excessive precision in the latitude, longitude, and location 
	fields is  nonsensical.  We would recommend no more than 
	four decimal places (0.000X) which represents a distance of approximately 
	36.4 feet. A 14 decimal number is simply a case of a great 
	precision but limited accuracy.

	\item The 11 address-related address fields (incident\_address, 
	street\_name, city, intersection\_street\_1 \& 2, cross\_street\_1 \& 2,  
	landmark, road\_ramp, bridge\_highway\_segment, and 
	taxi\_pick\_up\_location) should be put into standard USPS format. There 
	are several software packages which can normalize addresses to 
	these standards.

	\item There are multiple means by which missing data is 
	represented. These include ``UNKNOWN'' , ``N/A'', ``NA'', ``na'', and 
	a blank field, e.g. `` ''. Protocols should be established 
	regarding how to represent missing data.
\end{itemize}
